A White Noise Approach to Evolutionary Ecology

### 2020-04-12 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
including champagnat's unifying approach

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued Markov processes. This approach allows us to compute the full implications of a stochastic process such as demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    extra_dependencies: ["float","flafter"]
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Our goal in this manuscript is to develop a rigorous, but accessible approach to synthesize the stochastic dynamics of abundance, mean trait and heritable variation in biological populations for the study of theoretical evolutionary ecology. A primary aim of theoretical evolutionary ecology is the development of mathematical approaches to describe the evolution of populations and their interactions with both the biotic and abiotic environments in which they are embedded. Given this consideration, a natural scope for such an approach centers on quantifying the abundance dynamics of populations and the evolution of traits mediating their interactions as functions of relevant abiotic factors. Although taking into account abundance, phenotype and environment provides the basis for a partial understanding of the complex nature of biological communities, a deeper understanding must account for the effects of contemporary dispersal and the phylogeographic history of interacting lineages [@Kraft2007; @Hickerson2010; @Manceau2016; @markmcpeek2017] along with the genetic basis of ecologically relevant traits [@jeffreyconner2004; @FUSSMANN2007] and feedbacks between populations and the biogeochemical cycles they ultimately depend on [@michelloreau2010; @9781107648258]. It is therefore ideal that the development of any such mathematical approach anticipates extensions to account for these important factors shaping ecological communities, especially as empirical and conceptual work in these directions continues to grow [@AbdalaRoberts2014; @Klzsch2015; @Crutsinger2015; @Fitzpatrick2015; @Fitzpatrick2017; @Marx2017; @Rudman2017; @Skovmand2018; @VanNuland2019; @Harmon2019]. Furthermore, the approach would benefit from a stochastic component to capture the chance nature of biological reality [@Lande2003; @DeMeester2018; @Mubayi2019] and serve as a basis for the construction of statistical methods that measure evolutionary and ecological processes occuring in the wild. Such methods will tether theory to reality and allow for rigorous tests of hypotheses on the structure and behavior of ecological communities. In this paper we introduce a framework that establishes a formal connection between the continuous-time dynamics of abundance and quantitative traits in stochastically evolving populations. We then demonstrate the utility of our framework through the derivation and analysis of a model of diffuse coevolution and discuss how it can be extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which is intimately related to the approach we take here.

Reviewing these accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles. However, for populations with a continuum of types, such as a quantitative trait, this becomes a vexing mathematical challenge. Here we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. To show that our approach can be used to develop useful biological insights we derive a model of coevolution in an ecological network and use it to investigate the relationship between competition coefficients and selection gradients. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretic quantitative genetics. Observing that generalizing the PDE to a SPDE, we find a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by introducing a set of mathematical tools based on the calculus of white noise and discuss how a diffusion limit of a spatially structured branching process leads to the natural SPDE appropriate for our study. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. We employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find that the effect of random genetic drift on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

@@

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. Individuals are assumed to be haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include 1) mean fitness $\bar m$ determines the abundance dynamics of the entire population and 2) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

where $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and we have set $\kappa(x,y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties that we breifly take note of. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. By replacing $\kappa(x,y)=1$ in expression (\ref{special_m}) with a particular Gaussian function, we will find in \S\ref{coev} and SM \S\ref{diffuse} this fitness function becomes equivalent to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To clarify the distinction between this model and previous models of phenotypic evolution we refer to the PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work, to list just a few citations in this literature, has shown the importance of demographic stochasticity [@Turelli1980; @Schreiber2017] and evolutionary dynamics [@Case1986; @Schreiber2018] in determining coexistence of competing species. However, these studies consider evolution and demographic stochasticity separately. To our knowledge, there are no rigorous theoretical investigations of coexistence that account for evolutionary dynamics of quantitative characters and demographic stochasticity simultaneously. In fact, it appears that theoretical investigations combining these processes in a single species model to understand persistence have only just begun [@Gomulkiewicz2017]. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.

>>>>

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    extra_dependencies: ["float","flafter"]
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Our goal in this manuscript is to develop a rigorous, but accessible approach to synthesize the stochastic dynamics of abundance, mean trait and heritable variation in biological populations for the study of theoretical evolutionary ecology. A primary aim of theoretical evolutionary ecology is the development of mathematical approaches to describe the evolution of populations and their interactions with both the biotic and abiotic environments in which they are embedded. Given this consideration, a natural scope for such an approach centers on quantifying the abundance dynamics of populations and the evolution of traits mediating their interactions as functions of relevant abiotic factors. Although taking into account abundance, phenotype and environment provides the basis for a partial understanding of the complex nature of biological communities, a deeper understanding must account for the effects of dispersal and the phylogeographic history of interacting lineages [@Kraft2007; @Hickerson2010; @Manceau2016; @markmcpeek2017] along with the genetic basis of ecologically relevant traits [@jeffreyconner2004; @FUSSMANN2007] and feedbacks between populations and the biogeochemical cycles they ultimately depend on [@michelloreau2010; @9781107648258]. It is therefore ideal that the development of any such mathematical approach anticipates extensions to account for these important factors shaping ecological communities, especially as empirical and conceptual work in these directions continues to grow [@AbdalaRoberts2014; @Klzsch2015; @Crutsinger2015; @Fitzpatrick2015; @Fitzpatrick2017; @Marx2017; @Rudman2017; @Skovmand2018; @VanNuland2019; @Harmon2019]. Furthermore, the approach would benefit from a stochastic component to capture the chance nature of biological reality [@Lande2003; @DeMeester2018; @Mubayi2019] and serve as a basis for the construction of statistical methods that measure evolutionary and ecological processes occuring in the wild. Such methods will tether theory to reality and allow for rigorous tests of hypotheses on the structure and behavior of ecological communities. In this paper we introduce a framework that establishes a formal connection between the continuous-time dynamics of abundance and quantitative traits in stochastically evolving populations. We then demonstrate the utility of our framework through the derivation and analysis of a model of diffuse coevolution and discuss how it can be extended to account for the details mentioned above.

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which includes our approach as a special case. We note here there are two primary differences between our aims and those of Champagnat, Ferrière, and Méléard [-@Champagnat2006]. First, instead of developing a unifying approach to evolutionary theory, we focus on developing a stochastic synthesis of population dynamics and quantitative genetics including specific expressions for the dynamics of abundance, mean trait and additive genetic variance. Second, although Champagnat, Ferrière, and Méléard [-@Champagnat2006] provide rigorous proofs of results relevant to our approach, they do not communicate the nessecary tools we make use of for deriving the dynamics of abundance, mean trait and additive genetic variance. We aim to communicate these tools to a wide audience by providing an intuitive, yet accurate treatment of white noise and associated heuristics useful for deriving new stochastic models of evolutionary ecology.

Reviewing the above accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles [@Lande2009; @Parsons2010; @Gomulkiewicz2017]. However, for populations with a continuum of types, such as a quantitative trait, finding a formal approach to derive the evolutionary response to demographic stochasticity has remained a vexing mathematical challenge. Previous work integrating stochastic processes with quantitative genetic models have profited from the simplifying assumption that phenotypic and genetic variances remain constant over time, yeilding analytical approximations for the joint dynamics of abundance and the mean of a quantitative trait [@Engen2013]. These results not only ignore the evolution genetic variance, but they focus on environmental stochasticity and assume sufficiently large populations so that demographic stochasticity is negligible. Here we account for both demographic stochasticity and the evolution of additive genetic variance by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

Through accomplishing these goals this work yields five main results. Firstly, we develop a concise, but accurate approach for understanding SPDE and provide heuristics to empower mathematical evolutionary ecologists to derive novel models that relax key assumptions commonly made in combining models of evolutionary and abundance dynamics. Second, we formally derive general SDE that express the joint dynamics of abundance, mean trait and additive genetic variance for populations exhibiting a wide range of potential phenotypic distributions. Third, we simplify these SDE by approximating the abundance density of phenotypes with a Gaussian curve and integrate the results with a model of imperfect inheritance to find formula that generalize classical quantitative genetic expressions. This includes an expression for additive genetic variance analogous to the expression for the response in mean phenotype found by Lande [-@Lande1976]. Fourth, following classical niche theory we provide a rigorous mechanistic derivation of fitness for populations experiencing resource competition and abiotic stabilizing selection using diffusion limits of measure-valued branching processes. This calculation provides a demonstration for performing such diffusion limits that can be applied in other contexts. Fifth, we explore implications of the resulting model of diffuse coevolution by investigating the relationship between selection gradients and competition coefficients.

We now summarize the basic structure of the paper. To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretical quantitative genetics. Noting that a natural stochastic generalization of DAGA has been previously derived, we find a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by discussing how a diffusion limit of a spatially structured branching process (i.e., a measure-valued branching process) leads to the natural SPDE appropriate for our study. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. However, before we discuss measure-valued braching processes and their diffusion limits, we find it necessary to review some fundamental results on the calculus of white noise and introducing a set of mathematical tools based on these results. In \S\ref{wnc} and SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find that the effect of demographic stochasticity on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and strength of coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002] and Champagnat, Ferrière, and Méléard [-@Champagnat2006], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

@@

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include 1) mean fitness $\bar m$ determines the abundance dynamics of the entire population and 2) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

where $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and we have set $\kappa(x,y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties that we breifly take note of. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. By replacing $\kappa(x,y)=1$ in expression (\ref{special_m}) with a particular Gaussian function, we will find in \S\ref{coev} and SM \S\ref{diffuse} this fitness function becomes equivalent to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; and @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To clarify the distinction between this model and previous models of phenotypic evolution we refer to the PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work, to list just a few citations in this literature, has shown the importance of demographic stochasticity [@Turelli1980; @Schreiber2017] and evolutionary dynamics [@Case1986; @Schreiber2018] in determining coexistence of competing species. However, these studies consider evolution and demographic stochasticity separately. To our knowledge, there are no rigorous theoretical investigations of coexistence that account for evolutionary dynamics of quantitative characters and demographic stochasticity simultaneously and, in particular, no studies combining these processes that also include the evolution of additive genetic variance. In fact, only recently have rigorous theoretical investigations successfully combined these processes to understand persistence of a single population [@Gomulkiewicz2017]. Even so, these authors focused on a single species model with a finite number of types (i.e., a population genetic model). With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of quantitative characters and the effects of demographic stochasticity in a simple synthetic framework.
